A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-22 tidyindex.qmd
replace the reference with  the original articles

Linear combinations of variables are commonly used to reduce the dimension in statistical methodology, and chosen using a method like principal component analysis (PCA) or linear discriminant analysis [@james2013introduction; hastie2009elements], preparing contrasts to test particular elements in analysis of variance, or hand-crafted by a content-area expert. Linear combinations also form the basis for visualizing multivariate data, in methods such as tours [@wickham_tourr_2011].

>>>>

Linear combinations of variables are commonly used to reduce the dimension in statistical methodology, and chosen using a method like principal component analysis (PCA) [@hotelling1933analysis] or linear discriminant analysis [@fisher1936use], preparing contrasts to test particular elements in analysis of variance, or hand-crafted by a content-area expert. Linear combinations also form the basis for visualizing multivariate data, in methods such as tours [@wickham_tourr_2011].

### 2023-08-22 tidyindex.tex
replace the reference with  the original articles

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and variation in values. Here we discuss
a modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
pipeline framework usage: comparison of two different indexes designed
to monitor the spatio-temporal distribution of drought in Queensland,
Australia; the effect of dimension reduction choices on the Global
Gender Gap Index (GGGI) on countries' ranking; and how to calculate
bootstrap confidence intervals for the Standardized Precipitation Index
(SPI). The methods are supported by a new R package, called
\texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, sharp corners, frame hidden, breakable]}{\end{tcolorbox}}\fi

@@

Linear combinations of variables are commonly used to reduce the
dimension in statistical methodology, and chosen using a method like
principal component analysis (PCA) or linear discriminant analysis
{[}James et al. (2013); hastie2009elements{]}, preparing contrasts to
test particular elements in analysis of variance, or hand-crafted by a
content-area expert. Linear combinations also form the basis for
visualizing multivariate data, in methods such as tours (Wickham et al.
2011).

@@

\centering
\begin{tabular}{lr>{}l>{}r}
\toprule
\textbf{Variable} & \textbf{V-weight} & \textbf{Dimension} & \textbf{D-weight}\\
\midrule
Labour force participation & 0.199 & Economy & 0.25\\
Wage equality for similar work & 0.310 &  & \\
Estimated earned income & 0.221 &  & \\
Legislators senior officials and managers & 0.149 &  & \\
Professional and technical workers & 0.121 &  & \\
\hline
Literacy rate & 0.191 & Education & 0.25\\
Enrolment in primary education & 0.459 &  & \\
Enrolment in secondary education & 0.230 &  & \\
Enrolment in tertiary education & 0.121 &  & \\
\hline
Sex ratio at birth & 0.693 & Health & 0.25\\
Healthy life expectancy & 0.307 &  & \\
\hline
Women in parliament & 0.310 & Politics & 0.25\\
Women in ministerial positions & 0.247 &  & \\
Years with female head of state & 0.443 &  & \\
\bottomrule
\end{tabular}
\end{table}

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-james2013introduction}{}}%
James, Gareth, Daniela Witten, Trevor Hastie, Robert Tibshirani, et al.
2013. \emph{An Introduction to Statistical Learning}. Vol. 112.
Springer.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and variation in values. Here we discuss
a modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
pipeline framework usage: comparison of two different indexes designed
to monitor the spatio-temporal distribution of drought in Queensland,
Australia; the effect of dimension reduction choices on the Global
Gender Gap Index (GGGI) on countries' ranking; and how to calculate
bootstrap confidence intervals for the Standardized Precipitation Index
(SPI). The methods are supported by a new R package, called
\texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, frame hidden, sharp corners, breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

Linear combinations of variables are commonly used to reduce the
dimension in statistical methodology, and chosen using a method like
principal component analysis (PCA) (Hotelling 1933) or linear
discriminant analysis (Fisher 1936), preparing contrasts to test
particular elements in analysis of variance, or hand-crafted by a
content-area expert. Linear combinations also form the basis for
visualizing multivariate data, in methods such as tours (Wickham et al.
2011).

@@

\centering
\begin{tabular}{lrlr}
\toprule
\textbf{Variable} & \textbf{V-weight} & \textbf{Dimension} & \textbf{D-weight}\\
\midrule
Labour force participation & 0.199 & Economy & 0.25\\
Wage equality for similar work & 0.310 &  & \\
Estimated earned income & 0.221 &  & \\
Legislators senior officials and managers & 0.149 &  & \\
Professional and technical workers & 0.121 &  & \\
\hline
Literacy rate & 0.191 & Education & 0.25\\
Enrolment in primary education & 0.459 &  & \\
Enrolment in secondary education & 0.230 &  & \\
Enrolment in tertiary education & 0.121 &  & \\
\hline
Sex ratio at birth & 0.693 & Health & 0.25\\
Healthy life expectancy & 0.307 &  & \\
\hline
Women in parliament & 0.310 & Politics & 0.25\\
Women in ministerial positions & 0.247 &  & \\
Years with female head of state & 0.443 &  & \\
\bottomrule
\end{tabular}
\end{table}

@@

\leavevmode\vadjust pre{\hypertarget{ref-fisher1936use}{}}%
Fisher, Ronald A. 1936. {``The Use of Multiple Measurements in Taxonomic
Problems.''} \emph{Annals of Eugenics} 7 (2): 179--88.

@@

\leavevmode\vadjust pre{\hypertarget{ref-hotelling1933analysis}{}}%
Hotelling, Harold. 1933. {``Analysis of a Complex of Statistical
Variables into Principal Components.''} \emph{Journal of Educational
Psychology} 24 (6): 417.
